\ \\ \LARGE\bf Speculative Evaluation in Particle Swarm Optimization \thanks{Matthew Gardner, Andrew McNabb, and Kevin Seppi are with the Department of Computer Science, Brigham Young University, 3361 TMCB, Provo, UT 84602 (phone: 801-422-8717; email: \{mjg82,a,k\}@cs.byu.edu).}

### 2010-03-16 honors_thesis/journal_paper/paper.tex
Fixed some paragraphs, and changed the notation a bit
Made a lot of changes to the journal paper, from comments I got.

@@

The original PSO can be reproduced exactly, two iterations at a time, using
this technique.  However, if we relax the requirements of the algorithm, no
longer demanding that it strictly mimic the behavior of the original PSO, we
can introduce new speculative techniques that often out-perform PSO.

@@

At each iteration of the algorithm, the position $\ppos_t$ and velocity
$\pvel_t$ of each particle are updated as follows:
\begin{align}
\label{eq:velupdate}
	\pvel_{t+1} &=
		\constriction \left[ \pvel_t +
			\coeff_1\URand()\otimes(\pbest - \ppos_t) +
			\coeff_2\URand()\otimes(\nbest - \ppos_t)
		\right] \\
\label{eq:posupdate}
	\ppos_{t+1} &= \ppos_t + \pvel_{t+1}
\end{align}
where \( \URand() \) is a vector of random numbers drawn from a uniform
distribution, the \( \otimes \) operator is an element-wise vector
multiplication, $\pbest$ (called personal best) is the best position the
current particle has seen, and $\nbest$ (called neighborhood best) is the best
position the neighbors of the current particle have seen.  \( \coeff_1 \), \(
\coeff_2 \), and \( \constriction \) are parameters with prescribed values
required to ensure convergence (2.05, 2.05, and .73,
respectively)~\cite{clerc-tec02}.

@@

In some distributed frameworks, like the MapReduce framework that we used in
our experiments, rounds of message passing can be expensive.  The method just
described uses three rounds of message passing for every two iterations
(corresponding to steps 2, 5 and 10 in \alg{centralized}).  It is possible to
perform speculative evaluation in PSO with only one round of communication per
two iterations.  However, the methodology is tedius and is not the focus of
this paper, so we defer its description to the Appendix.

@@

We ran experiments to compare our speculative PSO algorithm to the standard
PSO algorithm.  At each iteration of the algorithms, we use one processor to
perform one function evaluation for one particle, be it speculative or
non-speculative.  The speculative algorithm actually performs two iterations of
PSO at each ``iteration,'' so we instead call each ``iteration'' a round of
evaluations.  For benchmark functions with very fast evaluations this may not
be the best use of parallelism in PSO.  But for many real-world applications,
the objective function takes on the order of minutes or more to evaluate; in
such cases our assumption is perfectly reasonable.

To use the same number of processors for each algorithm, speculative evaluation
needs a smaller swarm size than standard PSO.  For each of the topologies we
used, a particle has two neighbors other than itself.  As shown in
Table~\ref{tab:evals}, this results in $7$ speculative evaluations per
particle.  With one evaluation needed for the original, non-speculative
particle, we have $8p_s$ evaluations for every two iterations, where $p_s$ is
the number of particles in the speculative swarm.  The extra evaluations done
in speculative evaluation would instead be particles in regular PSO, so we
compare swarms of size $p_s$ in speculative evaluation with swarms of size
$8p_s$ in regular PSO.

@@

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{sphere.eps}
  \caption{Function Sphere with a swarm that uses 240 processors per round of
  evaluations.  Our speculative algorithm has a swarm of 30 particles and does
  two iterations per round, while standard PSO algorithms use 240 processors
  and do one iteration per round.}
  \label{fig:basic-sphere}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{griewank1.eps}
  \caption{Function Griewank with a swarm that uses 240 processors per round of
  evaluations.}
  \label{fig:basic-griewank1}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{griewank2.eps}
  \caption{Function Griewank with a swarm that uses 240 processors per round of
  evaluations.  Instead of showing average function value, we show the
  percentage of runs that have found the global optimum by each iteration.}
  \label{fig:basic-griewank2}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{griewank3.eps}
  \caption{Function Griewank with a swarm that uses 800 processors per round of
  evaluations.}
  \label{fig:basic-griewank3}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{rastrigin.eps}
  \caption{Function Rastrigin with a swarm that uses 240 processors per round
  of evaluations.}
  \label{fig:rastrigin}
\end{figure}

@@

@@

We show in \fig{sphere-pickbest} the results of using this modification on the
function Sphere.  For ease of labeling, we call the modified algorithm SEPSO
Pick Best.  As can be seen, where SEPSO did not manage to perform as well as
PSO Complete, SEPSO Pick Best handily outperforms it with the same number of
processors.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{sphere2.eps}
  \caption{Function Sphere with a swarm that uses 240 processors per round of
  evaluations, comparing SEPSO Pick Best with previous results.}
  \label{fig:sphere-pickbest}
\end{figure}

Sphere is a function for which very little exploration needs to be done, so
always keeping the best speculative position found makes intuitive sense.  A
less intuitive result occurs when using SEPSO Pick Best with the function
Griewank.  Griewank is a highly deceptive function prone to premature
convergence.  Our intuition on the behavior of picking the best speculative
child is that it would tend to prematurely converge more often than the
original PSO.  However, as seen in \fig{griewank-pickbest}, SEPSO Pick Best
increases the chance of finding the optimum while also reducing the number of
iterations required to converge.  Perhaps SEPSO Pick Best encourages local
exploitation, while the sparse communication of the Ring topology still allows
for global exploration.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{griewank4.eps}
  \caption{Function Sphere with a swarm that uses 240 processors per round of
  evaluations, comparing SEPSO Pick Best with previous results.}
  \label{fig:griewank-pickbest}
\end{figure}

\subsection{Pruning the Speculative Evaluations}

Because the problem facing speculative evaluation is having too many possible
evaluations to perform, a natural step to take is to eliminate some of the
evaluations.  If we could reliably predict which branch were going to be taken,
we could limit ourselves to one speculative evaluation per particle instead of
$2n+1$.  While we cannot with certainty predict the correct branch every time,
we can look at the statistics of the behavior of particles in general and
perhaps come up with a good guess, reducing the number of necessary speculative
evaluations.

\subsubsection{Branch Statistics}

@@

\begin{table}[h]
  \caption{Five Branches to Consider for Statistics}
  \label{tab:branches}
  \centering
  \begin{tabular}{rc|c}
	&$\pbest$ updated&Source of $\nbest$ update\\
	\hline
	1.&No update&No update\\
	2.&Updated&No update\\
	3.&Updated&The particle itself\\
	4.&No update&Some neighbor\\
	5.&Updated&Some neighbor\\
  \end{tabular}
\end{table}

\begin{table}[h]
  \caption{Branch Statistics in PSO}
  \label{tab:stats}
  \centering
  \begin{tabular}{c|c|c|c|c|c|c}
	Topology&Function&1&2&3&4&5\\
	\hline
	\hline
	Ring&Sphere&53.0\%&9.3\%&11.4\%&20.2\%&6.2\%\\
	&Griewank&51.7\%&8.4\%&12.2\%&20.7\%&7.0\%\\
	&Rastrigin&49.5\%&4.8\%&14.6\%&21.3\%&9.9\%\\
	&Rosenbrock&51.3\%&7.4\%&12.9\%&21.1\%&7.3\%\\
	\cline{2-7}
	&Average&{51.3\%}&{7.5\%}&{12.8\%}&{20.8\%}&
	{7.6\%}\\
	\hline
	\hline
	Random&Sphere&66.7\%&11.9\%&2.6\%&15.6\%&3.1\%\\
	&Griewank&69.0\%&10.9\%&2.5\%&14.9\%&2.7\%\\
	&Rastrigin&81.9\%&5.5\%&1.5\%&10.0\%&1.0\%\\
	&Rosenbrock&74.2\%&7.7\%&2.2\%&14.0\%&1.8\%\\
	\cline{2-7}
	&Average&{73.0\%}&{9.0\%}&{2.2\%}&{13.6\%}&
	{2.2\%}\\
	\hline
	\hline
	Complete&Sphere&31.9\%&9.2\%&0.2\%&45.1\%&13.5\%\\
	&Griewank&35.3\%&8.4\%&0.2\%&44.1\%&11.9\%\\
	&Rastrigin&47.7\%&6.7\%&0.2\%&38.2\%&7.0\%\\
	&Rosenbrock&35.3\%&3.4\%&0.3\%&54.4\%&6.6\%\\
	\cline{2-7}
	&Average&{37.6\%}&{6.9\%}&{.2\%}&{45.5\%}&
	{9.8\%}\\
	\hline
	\multicolumn{7}{l}{The numbers in the column headings refer to branch
	numbers in Table~\ref{tab:branches}.}
  \end{tabular}
\end{table}

@@

We now show with what percentage a particle takes each of these branches for
three different topologies and four different functions.  All of our statistics
are from swarms of 240 particles.  Brief experimentation showed that other
swarm sizes had similar statistics.  We ran 750 iterations on all combinations
of functions and topologies except for the functions Griewank and Rastrigin
with the Complete topology.  We found that those runs frequently converged past
machine precision after 500 iterations, and that led to erroneously high values
for the probability of stagnation.  Instead we ran for only 450 iterations on
those two combinations.  All of our results were averaged over 20 runs of the
algorithm; thus the probabilities presented are the averages of 3.6 million
trials for the branch taken (2.16 million for the two with only 450
iterations).  Table~\ref{tab:stats} contains the results, with the branch
number coming from Table~\ref{tab:branches}.  We have already defined the
functions Sphere, Griewank, and Rastrigin.  Rosenbrock is defined as
$f(\Vec{x}) = \sum_{i=1}^D \left(100(x_{i+1}-x_i^2)+(x_i-1)^2\right)$.

The probabilities presented in Table~\ref{tab:stats} are interesting in and of
themselves and could probably be used to better understand the characteristics
of various topologies.  It is notable that there is small variation between
functions in any given topology, but the variation across topologies is far
greater.  However, our concern is with speculative evaluation.  We are
interested in predicting the branch that any given particle will take at a
particular iteration.

@@

With the other topologies, however, there are a few branches with very high
probabilities, so we can have hope that cutting out some of the evaluations
that have low probability will lead to an increase in overall performance.  In
order to implement this kind of pruning, the only change that needs to be made
to \alg{centralized} is in Step 2, where the speculative children are
generated.  Simply generate a subset of the children that would normally be
generated.

\subsubsection{What if we guessed wrong?}

@@

One possibility is to leave the particle as it is, not accepting any of the
speculative evaluations, because none of them were correct.  This leads to
particles being at different iterations, as some particles guess correctly
while others do not.  Thus we lose exact compatibility with the original PSO,
though this particular relaxation is nothing new; asynchronous adaptations of
PSO do the same thing.  As an aside, it is equivalent in this case to simply
increment the iteration number of particles which fail to correctly predict
their branch.  This keeps the iteration number constant across all particles,
simplifying the work needed to be done in determining neighbors when dynamic
topologies are involved.

@@

In \fig{sphere-pruned}, we show the results of this simple pruning on the
function Sphere.  As can be seen, the larger number of particles that pruning
allows results in better results when picking the best speculative child.  
When rejecting all speculative children, which we with tongue in cheek call
Social Promotion, no real improvement is seen.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{sphere3.eps}
  \caption{Function Sphere with a swarm that uses 240 processors per round of
  evaluations, comparing pruning with previous results.}
  \label{fig:sphere-pruned}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{griewank5.eps}
  \caption{Function Griewank with a swarm that uses 240 processors per round of
  evaluations, comparing pruning with previous results.}
  \label{fig:griewank-pruned}
\end{figure}

\subsection{More than one iteration ahead}

@@

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{sphere4.eps}
  \caption{Function Sphere with a swarm that uses 240 processors per round of
  evaluations, comparing speculating several iterations ahead with previous
  results.}
  \label{fig:sphere-manyiters}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{griewank6.eps}
  \caption{Function Griewank with a swarm that uses 240 processors per round of
  evaluations, comparing speculating several iterations ahead with previous
  results.}
  \label{fig:griewank-manyiters}
\end{figure}

\subsection{Tokens}

@@

If we use the concept of tokens commonly used in multirobot stuff (see, for
example, \dots), we can specify a number of tokens available for particles to
use, where each token corresponds to a speculative evaluation.  Whichever
method is used to decide which evaluations to do, they are prioritized so that
particles with only one or two tokens perform the most valuable speculative
evaluations.  Particles that perform well keep their tokens, and particles that
perform poorly send one token (if they have one) to another random particle in
the swarm.

@@

There are thus two distinct uses for tokens.  First, they could be used to try
to evenly spread extra evaluations due to an incapatible number of particles
and processors.  In cursory experimentation, we found that this was not a very
effective use of tokens.  In virtually every case we tried, it was better to
simply change the number of particles to match the number of processors, and 
spread the speculative evaluations evenly throughout the swarm.

@@

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{griewank7.eps}
  \caption{Function Griewank with a swarm that uses 240 processors per round of
  evaluations, comparing token-based allocation with previous results.}
  \label{fig:griewank-token}
\end{figure}

@@

We have described a new technique for using processors in parallel PSO to
improve the performance of the algorithm.  To our knowledge, this is the first
time extra processors have been used to do anything in PSO besides increase the
swarm size.  In an increasingly parallel world, such advancements will prove to
be crucial in the continued effectiveness of PSO.

We have detailed how to perform speculative evaluation in PSO in several
different parallel frameworks.  Using this methodology, the behavior of the
original PSO algorithm can either be exactly reproduced, two iterations at a
time, or the behavior can be modified in order to improve performance.  We have
shown results that conclusively demonstrate the superiority of our techniques
over naively adding particles to the swarm when extra processors are available.

We have shown several different ways to modify the behavior of PSO when doing
speculative evaluation, and, as expected given the No Free Lunch Theorem,
different modifications show varied performance with different functions.
While our exploration of these techniques has been cursory in places, we
believe the results shown are sufficient to demonstrate their promise.

@@

We have speculated that our ``pick the best child'' method simply draws a
number of samples from the next iteration of the sampling distribution of PSO
and keeps one of them.  It would then be expected that the sampling
distribution of that method is very similar to the distribution of the original
PSO.  This remains to be proven, or even demonstrated empirically.

@@

We turned what used to be a very simple algorithm with only one major parameter
to tweak (the topology) into a big hairy mess with countless possibilities for
speculative evaluations.  Perhaps future research will be able to narrow down
the endless possibilities into the few best ones, making the algorithm simple
again.

@@

Large parallel clusters are often required to successfully optimize practical
modern problems.  To properly use PSO with such clusters, a balance needs to be
made between using processors to increase the swarm size and using them to
increase the speed of the algorithm, or for perhaps as yet unconsidered
techniques.  This work is a first step in that direction.

@@

\begin{figure}
  \centering
  \psfrag{pt-n}[C][C]{$\nonbest{\p}_{t}$}
  \psfrag{nt-n}[C][C]{$\nonbest{\nset}_{t}$}
  \psfrag{pt}[C][C]{$\p_{t}$}
  \includegraphics[width=.3\columnwidth]{messages1.eps}
  \caption{The production of $\p_{t}$ from the original particle
  $\nonbest{\p}_{t}$ and the messages $\nonbest{\nset}_{t}$.}
  \label{fig:messages1}
\end{figure}

>>>>

%Other math
\providecommand{\prob}{\ensuremath{\mathrm{Pr}}}

@@

The original PSO can be reproduced exactly, two iterations at a time, using
this technique.  However, if we relax the requirements of the algorithm, no
longer demanding that it strictly mimic the behavior of the original PSO, we
can introduce new speculative techniques that often out-perform PSO.  Some
modifications that we explore are picking the speculative position with the
best value instead of the position that matches the branch that PSO took;
pruning some of the speculative evaluations to increase the number of particles
in the swarm; and speculating several iterations ahead instead of just one.

@@

At each iteration of the algorithm, the position $\ppos_t$ and velocity
$\pvel_t$ of each particle are updated as follows:
\begin{align}
\label{eq:velupdate}
	\pvel_{t+1} &=
		\constriction \left[ \pvel_t +
			\coeff_1\URand()\otimes(\pbest - \ppos_t) +
			\coeff_2\URand()\otimes(\nbest - \ppos_t)
		\right] \\
\label{eq:posupdate}
	\ppos_{t+1} &= \ppos_t + \pvel_{t+1}
\end{align}
where \( \URand() \) is a vector of random numbers drawn from a uniform
distribution, the \( \otimes \) operator is an element-wise vector
multiplication, $\pbest$ (called personal best) is the best position the
current particle has seen, and $\nbest$ (called neighborhood best) is the best
position the neighbors of the current particle have seen~\cite{bratton-sis07}.
\( \coeff_1 \), \( \coeff_2 \), and \( \constriction \) are parameters with
prescribed values required to ensure convergence (2.05, 2.05, and .73,
respectively)~\cite{clerc-tec02}.

@@

In some distributed frameworks, like the MapReduce framework that we used in
our experiments, rounds of message passing can be expensive.  The method just
described uses three rounds of message passing for every two iterations
(corresponding to steps 2, 5 and 10 in \alg{centralized}).  It is possible to
perform speculative evaluation in PSO with only one round of communication per
two iterations.  However, the methodology is tedious and is not the focus of
this paper, so we defer its description to the Appendix.

@@

We ran experiments to compare our speculative PSO algorithm to the standard PSO
algorithm.  At each iteration of the algorithms, we use one processor to
perform one function evaluation for one particle, be it speculative or
non-speculative.  For benchmark functions with very fast evaluations this may
not be the best use of parallelism in PSO.  But for many real-world
applications, the objective function takes on the order of minutes or more to
evaluate; in such cases our choice of framework is reasonable.

The speculative algorithm actually performs two iterations of PSO at each
``iteration,'' so we instead call each ``iteration'' a round of evaluations.

To use the same number of processors (and hence evaluations) for each
algorithm, speculative evaluation needs a smaller swarm size than standard PSO.
For each of the topologies we used, a particle has three neighbors including
itself.  As shown in Table~\ref{tab:evals}, this results in $7$ speculative
evaluations per particle.  With one evaluation needed for the original,
non-speculative particle, we have $8p_s$ evaluations for every two iterations,
where $p_s$ is the number of particles in the speculative swarm.  The extra
evaluations done in speculative evaluation would instead be particles in
regular PSO, so we compare swarms of size $p_s$ in speculative evaluation with
swarms of size $8p_s$ in regular PSO.

@@

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{sphere}
  \caption{Function Sphere with a swarm that uses 240 processors per round of
  evaluations.  Our speculative algorithm has a swarm of 30 particles and does
  two iterations per round, while standard PSO algorithms use 240 processors
  and do one iteration per round.}
  \label{fig:basic-sphere}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{griewank1}
  \caption{Function Griewank with a swarm that uses 240 processors per round of
  evaluations.}
  \label{fig:basic-griewank1}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{griewank2}
  \caption{Function Griewank with a swarm that uses 240 processors per round of
  evaluations.  Instead of showing average function value, we show the
  percentage of runs that have found the global optimum by each iteration.}
  \label{fig:basic-griewank2}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{griewank3}
  \caption{Function Griewank with a swarm that uses 800 processors per round of
  evaluations.}
  \label{fig:basic-griewank3}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{rastrigin}
  \caption{Function Rastrigin with a swarm that uses 240 processors per round
  of evaluations.}
  \label{fig:rastrigin}
\end{figure}

@@

We outline three main improvements to speculative evaluation.  First, in
\sec{pickbest} we describe a method that uses all of the information found in
doing speculative evaluations.  Then Sections~\ref{sec:pruning}
through~\ref{sec:wrong} present a technique that reduces the number of
speculative evaluations that need to be done for each particle.  Finally,
\sec{manyiters} shows a method for speculating several iterations ahead,
instead of just one.  \sec{other} presents a few other ideas that seem
promising but need further development.

@@

We show in \fig{sphere-pickbest} the results of using this modification on the
function Sphere.  For ease of labeling, we call the modified algorithm Pick
Best.  As can be seen, where SEPSO did not manage to perform as well as PSO
Complete, Pick Best handily outperforms it with the same number of processors.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{sphere2}
  \caption{Function Sphere with a swarm that uses 240 processors per round of
  evaluations, comparing Pick Best with previous results.}
  \label{fig:sphere-pickbest}
\end{figure}

Sphere is a function for which very little exploration needs to be done, so
always keeping the best speculative position found makes intuitive sense.  A
less intuitive result occurs when using Pick Best with the function Griewank.
Griewank is a highly deceptive function prone to premature convergence.  Our
intuition on the behavior of picking the best speculative child is that it
would tend to prematurely converge more often than the original PSO.  However,
as seen in \fig{griewank-pickbest}, Pick Best increases the chance of finding
the optimum while also reducing the number of iterations required to converge.
Perhaps Pick Best encourages local exploitation, while the sparse communication
of the Ring topology still allows for global exploration.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{griewank4}
  \caption{Function Sphere with a swarm that uses 240 processors per round of
  evaluations, comparing Pick Best with previous results.}
  \label{fig:griewank-pickbest}
\end{figure}

\subsection{Pruning the Speculative Evaluations}
\label{sec:pruning}

Because the problem facing speculative evaluation is having too many possible
evaluations to perform, a natural step to take is to eliminate some of the
evaluations.  If we could reliably predict which branch were going to be taken,
we could limit ourselves to one speculative evaluation per particle instead of
$2n+1$.  As not all of the branches are evaluated in any given iteration, we
call this technique pruning.

A naive approach to pruning is to keep track of the last branch taken by each
particle and speculate on that branch.  This turns out to be correct far less
than half of the time on average.

A more principled approach would be to use the statistical behavior of PSO to
find probabilities of taking any particular branch.  While we cannot with
certainty predict which branch a particle will take every time, if we can use
statistics to narrow down the $2(n+1)$ possible evaluations to a few likely
candidates, we can decrease the amount of computation required to do
speculative evaluation and improve our performance.

\subsection{Branch Statistics}

@@

\begin{table}[ht]
  \caption{Five Branches to Consider for Statistics}
  \label{tab:branches}
  \centering
  \begin{tabular}{rc|c}
	&$\pbest$ updated&Source of $\nbest$ update\\
	\hline
	1.&No update&No update\\
	2.&Updated&No update\\
	3.&Updated&The particle itself\\
	4.&No update&Some neighbor\\
	5.&Updated&Some neighbor\\
  \end{tabular}
\end{table}

\begin{table}[ht]
  \caption{Branch Statistics in PSO}
  \label{tab:stats}
  \centering
  \begin{tabular}{c|c|c|c|c|c|c}
	Topology&Function&1&2&3&4&5\\
	\hline
	\hline
	Ring&Sphere&53.0\%&9.3\%&11.4\%&20.2\%&6.2\%\\
	&Griewank&51.7\%&8.4\%&12.2\%&20.7\%&7.0\%\\
	&Rastrigin&49.5\%&4.8\%&14.6\%&21.3\%&9.9\%\\
	&Rosenbrock&51.3\%&7.4\%&12.9\%&21.1\%&7.3\%\\
	\cline{2-7}
	&Average&{51.3\%}&{7.5\%}&{12.8\%}&{20.8\%}&
	{7.6\%}\\
	\hline
	\hline
	Random&Sphere&66.7\%&11.9\%&2.6\%&15.6\%&3.1\%\\
	&Griewank&69.0\%&10.9\%&2.5\%&14.9\%&2.7\%\\
	&Rastrigin&81.9\%&5.5\%&1.5\%&10.0\%&1.0\%\\
	&Rosenbrock&74.2\%&7.7\%&2.2\%&14.0\%&1.8\%\\
	\cline{2-7}
	&Average&{73.0\%}&{9.0\%}&{2.2\%}&{13.6\%}&
	{2.2\%}\\
	\hline
	\hline
	Complete&Sphere&31.9\%&9.2\%&0.2\%&45.1\%&13.5\%\\
	&Griewank&35.3\%&8.4\%&0.2\%&44.1\%&11.9\%\\
	&Rastrigin&47.7\%&6.7\%&0.2\%&38.2\%&7.0\%\\
	&Rosenbrock&35.3\%&3.4\%&0.3\%&54.4\%&6.6\%\\
	\cline{2-7}
	&Average&{37.6\%}&{6.9\%}&{.2\%}&{45.5\%}&
	{9.8\%}\\
	\hline
	\multicolumn{7}{l}{The numbers in the column headings refer to branch
	numbers in Table~\ref{tab:branches}.}
  \end{tabular}
\end{table}

@@

We seek to find the probability of taking any given branch, given whatever
information is needed: $\prob(B_t|\cdot)$.  In finding these probabilities, we
do not attempt to derive any distribution from the PSO equations, we simply
look at empirical distributions.  However, even with empirical distributions,
the problem with this approach is that it is not clear what information
influences the probability of taking a branch.  The failure of the naive
approach seems to say that the previous branch, $B_{t+1}$, is not incredibly
useful.  We look at two factors that we believe have a significant influence on
$\prob(B_t)$: topology and function.  Thus we are looking at $\prob(B_t|T,F)$.

We show in Table~\ref{tab:stats} with what percentage a particle takes each of
these branches for three different topologies and four different functions.
All of our statistics are from swarms of 240 particles.  Brief experimentation
showed that other swarm sizes had similar statistics.  We ran 750 iterations on
all combinations of functions and topologies except for the functions Griewank
and Rastrigin with the Complete topology.  We found that those runs frequently
converged past machine precision after 500 iterations, and that led to
erroneously high values for the probability of stagnation.  Instead we ran for
only 450 iterations on those two combinations.  All of our results were
averaged over 20 runs of the algorithm; thus the probabilities presented are
the averages of 3.6 million trials for the branch taken (2.16 million for the
two with only 450 iterations).  Table~\ref{tab:stats} contains the results,
with the branch number coming from Table~\ref{tab:branches}.  We have already
defined the functions Sphere, Griewank, and Rastrigin.  Rosenbrock is defined
as $f(\Vec{x}) = \sum_{i=1}^D \left(100(x_{i+1}-x_i^2)+(x_i-1)^2\right)$.

The probabilities presented in Table~\ref{tab:stats} are interesting in and of
themselves and could probably be used to better understand the characteristics
of various topologies.  It is notable that there is small variation between
functions in any given topology, but the variation across topologies is far
greater.  However, our concern is with speculative evaluation.  We are
interested in predicting the branch that any given particle will take at a
particular iteration.  For our purposes, it appears that given a topology, the
probability of selecting a branch and the function are independent, or
$\prob(B_t|T,F) \approx \prob(B_t|T)$.

@@

With the other topologies, however, there are a few branches with very high
probabilities, so we can have hope that cutting out some of the evaluations
that have low probability will lead to an increase in overall performance.  In
order to implement this kind of pruning, the only change that needs to be made
to \alg{centralized} is in Step 2, where the speculative children are
generated.  Instead of generating all possible speculative children, generate
the subset of the children that is desired.

\subsection{What if we guessed wrong?}
\label{sec:wrong}

@@

One possibility is to leave the particle as it is, not accepting any of the
speculative evaluations, because none of them were correct.  This leads to
particles being at different iterations, as some particles guess correctly
while others do not.  Thus we lose exact compatibility with the original PSO,
though this particular relaxation is nothing new; asynchronous adaptations of
PSO do the same thing.  As an aside, it is equivalent in this case to simply
increment the iteration number of particles which fail to correctly predict
their branch.  This keeps the iteration number constant across all particles,
simplifying the work needed to be done in determining neighbors when dynamic
topologies are involved.  We call this technique Social Promotion.

@@

In \fig{sphere-pruned}, we show the results of this simple pruning on the
function Sphere.  As can be seen, the larger number of particles that pruning
allows results in better results when picking the best speculative child.  
When rejecting all speculative children no real improvement is seen.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{sphere3}
  \caption{Function Sphere with a swarm that uses 240 processors per round of
  evaluations, comparing pruning with previous results.}
  \label{fig:sphere-pruned}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{griewank5}
  \caption{Function Griewank with a swarm that uses 240 processors per round of
  evaluations, comparing pruning with previous results.}
  \label{fig:griewank-pruned}
\end{figure}

\subsection{More than one iteration ahead}
\label{sec:manyiters}

@@

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{sphere4}
  \caption{Function Sphere with a swarm that uses 240 processors per round of
  evaluations, comparing speculating several iterations ahead with previous
  results.}
  \label{fig:sphere-manyiters}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{griewank6}
  \caption{Function Griewank with a swarm that uses 240 processors per round of
  evaluations, comparing speculating several iterations ahead with previous
  results.}
  \label{fig:griewank-manyiters}
\end{figure}

\subsection{Other possibilities}
\label{sec:other}

@@

If we use the concept of tokens commonly used in multi-robot stuff (see, for
example, \dots), we can specify a number of tokens available for particles to
use, where each token corresponds to a speculative evaluation.  Whichever
method is used to decide which evaluations to do, they are prioritized so that
particles with only one or two tokens perform the most valuable speculative
evaluations.  Particles that perform well keep their tokens, and particles that
perform poorly send one token (if they have one) to another random particle in
the swarm.

@@

There are thus two distinct uses for tokens.  First, they could be used to try
to evenly spread extra evaluations due to an incompatible number of particles
and processors.  In cursory experimentation, we found that this was not a very
effective use of tokens.  In virtually every case we tried, it was better to
simply change the number of particles to match the number of processors, and 
spread the speculative evaluations evenly throughout the swarm.

@@

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{griewank7}
  \caption{Function Griewank with a swarm that uses 240 processors per round of
  evaluations, comparing token-based allocation with previous results.}
  \label{fig:griewank-token}
\end{figure}

@@

We have described a new technique for using processors in parallel PSO to
improve the performance of the algorithm.  To our knowledge, this is the first
time extra processors have been used to do anything in PSO besides increase the
swarm size.  In an increasingly parallel world, such advancements will prove to
be crucial to the continued effectiveness of PSO.

We have detailed how to perform speculative evaluation in PSO in several
different parallel frameworks.  Using this methodology, the behavior of the
original PSO algorithm can either be exactly reproduced, two iterations at a
time, or the behavior can be modified in order to improve performance.  While
exactly reproducing PSO often uses too many extra processors to be useful, when
we allow ourselves some freedom with the algorithm we see great improvements
over previous methods.  We have shown results that conclusively demonstrate the
superiority of our techniques for several functions over naively adding
particles to the swarm when extra processors are available.

Though our methods show great improvements on some functions, they do not work
for all functions.  As is commonly known, in PSO there is a trade-off between
exploration and exploitation.  Some functions need only minimal exploration,
and some never seem to have enough.  Increasing the swarm size is a natural way
to increase exploration in a parallel environment.  However, once ``enough''
exploration has been reached for any particular function, adding additional
particles adds only incremental benefits.  A better use of the additional
processors, as we have shown, is to perform some amount of speculative
evaluation.

Sphere is a function for which only a very small amount of exploration needs to
be done, and our speculative methods show improvements on Sphere of over 40
orders of magnitude.  Such functions are perfect for our methods.

Rastrigin, on the other hand, is a function for which increasing the swarm size
up to 4000 particles still shows improvements in the performance of the
algorithm.  With such functions, the smaller swarm size required by speculative
evaluation is not able to produce enough exploration to perform better than
standard PSO at the swarm sizes we were able to experiment with.

Griewank is a function in between Sphere and Rastrigin.  It is highly deceptive
and is prone to premature convergence, but by adding particles to the swarm a
point is reached where ``enough'' exploration is done, and the algorithm finds
the optimum essentially all of the time.  For such functions, the best approach
seems to be to increase the swarm size until ``enough'' exploration is reached,
then use extra processors to perform speculative evaluation and increase the
number of iterations performed.  Sphere and Rastrigin can be thought of as
special cases of these types of functions; Sphere simply needs a very small
swarm size to produce ``enough'' exploration, and Rastrigin requires a very
large swarm.  We expect that for all functions there is a swarm size for which
additional particles are less useful than additional iterations.

Large parallel clusters are often required to successfully optimize practical
modern problems.  To properly use PSO with such clusters, a balance needs to be
made between using processors to increase the swarm size and using them to
increase the speed of the algorithm.  This work is a first step in that
direction.

@@

We have speculated that our Pick Best method simply draws a number of samples
from the next iteration of the sampling distribution of PSO and keeps one of
them.  It would then be expected that the sampling distribution of that method
is very similar to the distribution of the original PSO.  This remains to be
proven, or even demonstrated empirically.

@@

We turned what used to be a very simple algorithm with only one major parameter
to tweak (the topology) into a convoluted mess with countless possibilities for
speculative evaluations.  Perhaps future research will be able to narrow down
the endless possibilities into the few best ones, making the algorithm simple
again.  We have presented some small amount of exploration of the possible
alternatives, but our discussion is by no means exhaustive or conclusive.

@@

@@

\begin{figure}
  \centering
  \psfrag{pt-n}[C][C]{$\nonbest{\p}_{t}$}
  \psfrag{nt-n}[C][C]{$\nonbest{\nset}_{t}$}
  \psfrag{pt}[C][C]{$\p_{t}$}
  \includegraphics[width=.3\columnwidth]{messages1}
  \caption{The production of $\p_{t}$ from the original particle
  $\nonbest{\p}_{t}$ and the messages $\nonbest{\nset}_{t}$.}
  \label{fig:messages1}
\end{figure}
